Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-08-10 ms_pre_post.Rmd
Model section edits

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance strategies (hereafter simply 'inbreeding strategies' because the model did not preclude evolution of inbreeding preference), in a small focal population of females and males by tracking the dynamics of alleles underlying reproductive strategies expressed by females. We thereby track evolutionary dynamics given internally consistent patterns of relatedness caused by non-random mating and capture effects of mutation, gene flow, drift, and selection.

Complex traits such as reproductive strategies are likely to be polygenic [@Evans2008]. Hence, we model individuals with 10 physically unlinked diploid loci (and therefore 20 alleles), underlying each of three reproductive strategy traits: tendency for polyandry ($P_{a}$, i.e., 'polyandry' alleles), pre-copulatory inbreeding strategy ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding strategy ($F_{a}$, i.e., 'fertilisation' alleles). All individuals therefore have 30 diploid loci (and hence 60 alleles) in total, each of which can take the value of any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014].

@@

In contrast, individuals' phenotypic values for tendency for polyandry ($P_{p}$) cannot map directly onto their genotypic values ($P_{g}$) because $P_{g}$ can evolve to be negative, but females cannot mate with a negative number of additional males [e.g., @Shuker2007; @Evans2013]. Rather, polyandry is more realistically considered as a "threshold trait", whereby continuous genotypic variation translates into expression of discrete phenotypic value(s) at some threshold [@Lynch1998; @Roff1996; @Roff1998; @Duthie]. Accordingly, we allow individuals' phenotypic values for tendency for polyandry to equal genotypic values ($P_{p} = P_{g}$) if $P_{g} \geq 0$, but set $P_{p} = 0$ if $P_{g} < 0$. A negative $P_{g}$ value therefore generates a female that is phenotypically monandrous, while a positive $P_{g}$ value generates a female that can express some degree of polyandry (see details below). Polyandry is therefore influenced by continuous genetic variation but only expressed when $P_{g} > 0$. One important general property of such threshold traits is that deleterious traits are occasionally expressed despite sustained negative selection because recombination among deleterious alleles can cause the underlying genotypic value to exceed the threshold for expression [@Roff1996; @Roff1998].

@@

Phenotypic values of the three reproductive strategy traits each incur set costs that combine to independently increase the probability that a focal female will die before mating. Numerous forms and mechanisms of direct costs of reproductive strategies could potentially be hypothesised and modelled; the most appropriate formulation depends on the question posed (see Discussion). By allowing costs of the three strategies to be directly and independently controlled, our model facilitates direct comparison of evolution of pre-copulatory versus post-copulatory inbreeding avoidance given known cost asymmetries. Such costs on female survival probability are biologically reasonable. For example, polyandrous females that undertake increased mate searching or mating can experience increased predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice can increase the risk of mortality due to harm caused by sexual conflict over mating, and also risk complete mating failure [which equates to pre-reproductive mortality in semelparous organisms; @Rowe1994; @Kokko2013]. Finally, females that express post-copulatory choice can pay up-front energetic costs, which might result in trade-offs with survival due to developing physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014].

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. It also assumes that inbreeding does not covary with inbreeding load (i.e., no purging). This formulation ensures that the relationship between $k_{ij}$ and the magnitude of inbreeding depression in offspring is consistent across replicate simulations and is further justified because previous genetically-explicit modelling [@Duthie2016a] showed that inbreeding avoidance in biparental populations has a negligible effect on load given small-effect deleterious mutations [see also @Wang1999; @Guillaume2006]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection) so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Offspring that are viable after the Bernoulli trial become adults. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, not where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a] or outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are not precluded from evolving, and could potentially arise due to mutation or drift.

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where pre-copulatory inbreeding strategy was cost free ($c_{M}=0$) but post-copulatory inbreeding strategy was moderately costly ($c_{F}=0.02$), and vice versa. We compare evolutionary trajectories with those of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overwhelming direct negative selection on polyandry [@Duthie]. This value is therefore appropriate to illustrate the different evolutionary consequences that could result from asymmetrical costs, but simulations with additional cost value combinations are provided in Supporting Information.

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance, and resulting evolution, is affected by the other strategy of inbreeding avoidance already being fixed in the population, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and of post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., $F_{p}$ and $P_{p}$ were expressed but did not evolve further). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

>>>>

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance strategies (hereafter simply 'inbreeding strategies' because the model did not preclude evolution of inbreeding preference), in a small focal population by tracking the dynamics of alleles underlying reproductive strategies expressed by females. We thereby track evolutionary dynamics given internally consistent patterns of relatedness caused by non-random mating and capture effects of mutation, gene flow, drift, and selection.

Complex traits such as reproductive strategies are likely to be polygenic [@Evans2008]. Hence, we model individuals with 10 physically unlinked diploid loci (i.e., 20 alleles), underlying each of three reproductive strategy traits: tendency for polyandry ($P_{a}$, 'polyandry' alleles), pre-copulatory inbreeding strategy ($M_{a}$, 'mating' alleles), and post-copulatory inbreeding strategy ($F_{a}$, 'fertilisation' alleles). All individuals therefore have 30 diploid loci (i.e., 60 alleles) in total, each of which can take the value of any real number [continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014].

@@

In contrast, individuals' phenotypic values for tendency for polyandry ($P_{p}$) cannot map directly onto their genotypic values ($P_{g}$) because $P_{g}$ can evolve to be negative, but females cannot mate with a negative number of additional males [e.g., @Shuker2007; @Evans2013]. Rather, we considered polyandry "threshold trait", whereby continuous genotypic variation translates into expression of discrete phenotypic value(s) at some threshold [@Lynch1998; @Roff1996; @Roff1998; @Duthie]. Accordingly, we allow individuals' phenotypic values for tendency for polyandry to equal genotypic values ($P_{p} = P_{g}$) if $P_{g} \geq 0$, but set $P_{p} = 0$ if $P_{g} < 0$. A negative $P_{g}$ value therefore generates a female that is phenotypically monandrous, while a positive $P_{g}$ value generates a female that can express some degree of polyandry (see details below). Polyandry is therefore influenced by continuous genetic variation but only expressed when $P_{g} > 0$. One important general property of such threshold traits is that deleterious traits are occasionally expressed despite sustained negative selection because recombination among deleterious alleles can cause the underlying genotypic value to exceed the threshold for expression [@Roff1996; @Roff1998].

@@

Phenotypic values of the three reproductive strategy traits each incur set costs that combine to independently increase the probability that a focal female will die before mating. Numerous forms and mechanisms of direct costs of reproductive strategies could be hypothesised and modelled; the most appropriate formulation depends on the question (see Discussion). By allowing costs of the three strategies to be directly and independently controlled, our model facilitates direct comparison of evolution of pre-copulatory versus post-copulatory inbreeding avoidance given known cost asymmetries. Qualitatively, such costs on female survival probability are biologically reasonable. For example, polyandrous females that undertake increased mate searching or mating can experience increased predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice can increase the risk of mortality due to harm caused by sexual conflict over mating, and also risk complete mating failure [which equates to pre-reproductive mortality in semelparous organisms; @Rowe1994; @Kokko2013]. Finally, females that express post-copulatory choice can pay up-front energetic costs, which might result in trade-offs with survival due to developing physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014].

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. It also assumes that inbreeding does not covary with inbreeding load (i.e., no purging). This formulation ensures that the relationship between $k_{ij}$ and the magnitude of inbreeding depression in offspring is consistent across replicate simulations. This choice, as opposed to a more mechanistic model of inbreeding depression that allows purging, is further justified because previous genetically-explicit modelling [@Duthie2016a] showed that inbreeding avoidance in biparental populations has a negligible effect on load given small-effect deleterious mutations [see also @Wang1999; @Guillaume2006].

We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection) so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Offspring that are viable after the Bernoulli trial become adults. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, not where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a] or outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are not precluded from evolving, and could potentially arise due to mutation or drift.

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where pre-copulatory inbreeding strategy was cost free ($c_{M}=0$) but post-copulatory inbreeding strategy was moderately costly ($c_{F}=0.02$), and vice versa. We compare evolutionary trajectories with those of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overwhelming direct negative selection on polyandry [@Duthie]. This value is therefore appropriate to illustrate the different evolutionary consequences that could result from asymmetrical costs. Results from simulations with additional cost value combinations are provided in Supporting Information.

To address how selection on an initially rare inbreeding avoidance strategy, and resulting evolution, is affected by the other strategy of inbreeding avoidance already being fixed in the population, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and of post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., $F_{p}$ and $P_{p}$ were expressed but did not evolve further). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.
